---
title: "HRS Allostatic (over)Load Dataset Construction"
subtitle: "CHORDS Lab – Washington State University"
author:
- "David Rice"
- "Jason Cross"
- "Shawna Beese"
date: "Last updated: 2026-02-21"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Overview

This R Markdown script walks through the construction of a dataset for allostatic (over)load research using data from the [Health and Retirement Study (HRS)](https://hrs.isr.umich.edu/).

⚠️ Before continuing, make sure you’ve downloaded and organized the required raw data. See [`HRS_access_guide.md`](https://github.com/CHORDSLab/Allostatic-over-Load-Repository/blob/main/HRS/HRS_access_guide.md) for instructions.  

The [`HRS_access_guide.md`](https://github.com/CHORDSLab/Allostatic-over-Load-Repository/blob/main/HRS/HRS_access_guide.md) will walk you through downloading the necessary public and restricted-access datasets and arranging them in a folder structure compatible with this script.

---

### What You’ll Need to Run This Script

- Raw HRS data, organized as described in the guide
- R environment with the following packages: `SAScii`, `dplyr`, `stringr`, `Hmisc`
- Familiarity with HRS data formats and .sas scripts

💡 Reminder: Access to biomarker files requires registration and approval from HRS. Passphrases for restricted datasets are provided by HRS via email upon approval.

## Setup & Libraries

First, make sure you’ve installed the required packages. If you haven’t installed them yet, run the following commands in your console:

```{r install-packages, eval=FALSE}
# install.packages("SAScii")
# install.packages("dplyr")
# install.packages("stringr")
# install.packages("Hmisc")
```

Once installed, load the required libraries:

```{r libraries}
library(SAScii)
library(dplyr)
library(stringr)
library(Hmisc)
```

---

## 📁 Set File Path

```{r file-path}
# Set to the location of your "HRS Data Products" folder
path <- "PATH/TO/YOUR/HRS Data Products" # Replace with the full path to your local "HRS Data Products" folder
```

---

## Part 1 – Load, Clean, and Combine HRS Waves (2006-2016)

In this section, we’ll load HRS waves from 2006-2016, clean key variables, and combine them into a unified dataset.

---

### Load 2016 HRS Core + Tracker Data

```{r}

#####################
# Load 2016 HRS Core + Tracker Data
#####################

# Set path to 2016 HRS Core folder
setwd(file.path(path, "HRS Core Data/h16core"))

# Load datasets
preload_2016 <- read.SAScii("h16da/H16PR_R.da", "h16sas/H16PR_R.sas")
demo_2016    <- read.SAScii("h16da/H16B_R.da",  "h16sas/H16B_R.sas")
meas_2016    <- read.SAScii("h16da/H16I_R.da",  "h16sas/H16I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2016 <- add_id(preload_2016)
demo_2016    <- add_id(demo_2016)
meas_2016    <- add_id(meas_2016)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2016 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2016) from Cross-Wave Tracker File
    WT_RESP = PWGTR,     # 2016 weight: respondent level
    WT_PM   = PPMWGTR,   # 2016 physical measures subsample weight
    WT_BIO  = PBIOWGTR,  # 2016 biomarker subsample weight
    WT_LBQ  = PLBWGTR,   # 2016 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2016 <- preload_2016 %>%
  select(unique_id, sex = PX060_R, birth_year = PX067_R) %>%
  left_join(
    demo_2016 %>% select(unique_id, race = PB089M1M, ethnicity = PB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2016 %>% select(unique_id, pulse1 = PI861, pulse2 = PI866, pulse3 = PI871,
                         waist = PI907, height = PI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2016, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2016)

# Preview the joined 2016 Core + Tracker data
glimpse(joined_core_2016)

```

---

### Load 2014 HRS Core + Tracker Data

```{r}

#####################
# Load 2014 HRS Core + Tracker Data
#####################

# Set path to 2014 HRS Core folder
setwd(file.path(path, "HRS Core Data/h14core"))

# Load datasets
preload_2014 <- read.SAScii("h14da/H14PR_R.da", "h14sas/H14PR_R.sas")
demo_2014    <- read.SAScii("h14da/H14B_R.da",  "h14sas/H14B_R.sas")
meas_2014    <- read.SAScii("h14da/H14I_R.da",  "h14sas/H14I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2014 <- add_id(preload_2014)
demo_2014    <- add_id(demo_2014)
meas_2014    <- add_id(meas_2014)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2014 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2014) from Cross-Wave Tracker File
    WT_RESP = OWGTR,     # 2014 weight: respondent level
    WT_PM   = OPMWGTR,   # 2014 physical measures subsample weight
    WT_BIO  = OBIOWGTR,  # 2014 biomarker subsample weight
    WT_LBQ  = OLBWGTR,   # 2014 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2014 <- preload_2014 %>%
  select(unique_id, sex = OX060_R, birth_year = OX067_R) %>%
  left_join(
    demo_2014 %>% select(unique_id, race = OB089M1M, ethnicity = OB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2014 %>% select(unique_id, pulse1 = OI861, pulse2 = OI866, pulse3 = OI871,
                         waist = OI907, height = OI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2014, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2014)

# Preview the joined 2014 Core + Tracker data
glimpse(joined_core_2014)

```

---

### Load 2012 HRS Core + Tracker Data

```{r}

#####################
# Load 2012 HRS Core + Tracker Data
#####################

# Set path to 2012 HRS Core folder
setwd(file.path(path, "HRS Core Data/h12core"))

# Load datasets
preload_2012 <- read.SAScii("h12da/H12PR_R.da", "h12sas/H12PR_R.sas")
demo_2012    <- read.SAScii("h12da/H12B_R.da",  "h12sas/H12B_R.sas")
meas_2012    <- read.SAScii("h12da/H12I_R.da",  "h12sas/H12I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2012 <- add_id(preload_2012)
demo_2012    <- add_id(demo_2012)
meas_2012    <- add_id(meas_2012)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2012 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2012) from Cross-Wave Tracker File
    WT_RESP = NWGTR,     # 2012 weight: respondent level
    WT_PM   = NPMWGTR,   # 2012 physical measures subsample weight
    WT_BIO  = NBIOWGTR,  # 2012 biomarker subsample weight
    WT_LBQ  = NLBWGTR,   # 2012 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2012 <- preload_2012 %>%
  select(unique_id, sex = NX060_R, birth_year = NX067_R) %>%
  left_join(
    demo_2012 %>% select(unique_id, race = NB089M1M, ethnicity = NB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2012 %>% select(unique_id, pulse1 = NI861, pulse2 = NI866, pulse3 = NI871,
                         waist = NI907, height = NI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2012, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2012)

# Preview the joined 2012 Core + Tracker data
glimpse(joined_core_2012)

```

---

### Load 2010 HRS Core + Tracker Data

```{r}

#####################
# Load 2010 HRS Core Data
#####################

# Set path to 2010 HRS Core folder
setwd(file.path(path, "HRS Core Data/h10core"))

# Load datasets
preload_2010 <- read.SAScii("h10da/H10PR_R.da", "h10sas/H10PR_R.sas")
demo_2010    <- read.SAScii("h10da/H10B_R.da",  "h10sas/H10B_R.sas")
meas_2010    <- read.SAScii("h10da/H10I_R.da",  "h10sas/H10I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2010 <- add_id(preload_2010)
demo_2010    <- add_id(demo_2010)
meas_2010    <- add_id(meas_2010)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2010 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2010) from Cross-Wave Tracker File
    WT_RESP = MWGTR,     # 2010 weight: respondent level
    WT_PM   = MPMWGTR,   # 2010 physical measures subsample weight
    WT_BIO  = MBIOWGTR,  # 2010 biomarker subsample weight
    WT_LBQ  = MLBWGTR,   # 2010 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2010 <- preload_2010 %>%
  select(unique_id, sex = MX060_R, birth_year = MX067_R) %>%
  left_join(
    demo_2010 %>% select(unique_id, race = MB089M1M, ethnicity = MB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2010 %>% select(unique_id, pulse1 = MI861, pulse2 = MI866, pulse3 = MI871,
                         waist = MI907, height = MI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2010, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2010)

# Preview the joined 2010 Core + Tracker data
glimpse(joined_core_2010)

```

---

### Load 2008 HRS Core + Tracker Data

```{r}

#####################
# Load 2008 HRS Core Data
#####################

# Set path to 2008 HRS Core folder
setwd(file.path(path, "HRS Core Data/h08core"))

# Load datasets
preload_2008 <- read.SAScii("h08da/H08PR_R.da", "h08sas/H08PR_R.sas")
demo_2008    <- read.SAScii("h08da/H08B_R.da",  "h08sas/H08B_R.sas")
meas_2008    <- read.SAScii("h08da/H08I_R.da",  "h08sas/H08I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2008 <- add_id(preload_2008)
demo_2008    <- add_id(demo_2008)
meas_2008    <- add_id(meas_2008)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2008 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2008) from Cross-Wave Tracker File
    WT_RESP = LWGTR,     # 2008 weight: respondent level
    WT_PM   = LPMWGTR,   # 2008 physical measures subsample weight
    WT_BIO  = LBIOWGTR,  # 2008 biomarker subsample weight
    WT_LBQ  = LLBWGTR,   # 2008 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2008 <- preload_2008 %>%
  select(unique_id, sex = LX060_R, birth_year = LX067_R) %>%
  left_join(
    demo_2008 %>% select(unique_id, race = LB089M1M, ethnicity = LB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2008 %>% select(unique_id, pulse1 = LI861, pulse2 = LI866, pulse3 = LI871,
                         waist = LI907, height = LI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2008, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2008)

# Preview the joined 2008 Core + Tracker data
glimpse(joined_core_2008)

```

---

### Load 2006 HRS Core + Tracker Data

```{r}

#####################
# Load 2006 HRS Core Data
#####################

# Set path to 2006 HRS Core folder
setwd(file.path(path, "HRS Core Data/h06core"))

# Load datasets
preload_2006 <- read.SAScii("h06da/H06PR_R.da", "h06sas/H06PR_R.sas")
demo_2006    <- read.SAScii("h06da/H06B_R.da",  "h06sas/H06B_R.sas")
meas_2006    <- read.SAScii("h06da/H06I_R.da",  "h06sas/H06I_R.sas")

# Add unique identifier
add_id <- function(df) df %>% mutate(unique_id = paste(HHID, PN, sep = "_"))
preload_2006 <- add_id(preload_2006)
demo_2006    <- add_id(demo_2006)
meas_2006    <- add_id(meas_2006)

# Load weights + design vars from Cross-Wave Tracker File
tracker_2006 <- read.SAScii(
  file.path(path, "trk2022v1/TRK2022TR_R.da"),
  file.path(path, "trk2022v1/TRK2022TR_R.sas")
) %>%
  add_id() %>%
  transmute(
    unique_id,

    # module-specific weights (2006) from Cross-Wave Tracker File
    WT_RESP = KWGTR,     # 2006 weight: respondent level
    WT_PM   = KPMWGTR,   # 2006 physical measures subsample weight
    WT_BIO  = KBIOWGTR,  # 2006 biomarker subsample weight
    WT_LBQ  = KLBWGTR,   # 2006 leave-behind questionnaire subsample weight

    # design variables
    STRATUM = STRATUM,   # Stratum ID
    SECU    = SECU       # Sampling Error Computation Unit
  )

# Join and select relevant variables
joined_core_2006 <- preload_2006 %>%
  select(unique_id, sex = KX060_R, birth_year = KX067_R) %>%
  left_join(
    demo_2006 %>% select(unique_id, race = KB089M1M, ethnicity = KB029M1M),
    by = "unique_id"
  ) %>%
  left_join(
    meas_2006 %>% select(unique_id, pulse1 = KI861, pulse2 = KI866, pulse3 = KI871,
                         waist = KI907, height = KI834),
    by = "unique_id"
  ) %>%
  left_join(tracker_2006, by = "unique_id") %>%   # <-- weights/design vars
  mutate(survey_year = 2006)

# Preview the joined 2006 Core + Tracker data
glimpse(joined_core_2006)

```

---

### Combine 2006-2016 HRS Core + Tracker Data

```{r}

#Join all years by stacking 
full_core = rbind(joined_core_2016, joined_core_2014, joined_core_2012, joined_core_2010, joined_core_2008, joined_core_2006)

# Preview Full Core + Tracker data
glimpse(full_core)

```

---

## Loading In Biomarker Data

---

### Load 2016 HRS Biomarker Data

```{r}

#####################
# Load 2016 HRS Biomarker Data
#####################

# Set path to 2016 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/BIOMK16BL"))

# Load biomarker data
bio_2016 <- read.SAScii("BIOMK16BL_R.da", "BIOMK16BL_R.sas")

# Add unique ID and select relevant variables
bio_2016_trim <- bio_2016 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(
    unique_id,
    A1C = PA1C_ADJ,
    HDL = PHDL_ADJ,
    CRP = PCRP_ADJ
  ) %>%
  mutate(survey_year = 2016)

# Preview 2016 Biomarker Data
glimpse(bio_2016_trim)

```

---

### Load 2014 HRS Biomarker Data

```{r}

#####################
# Load 2014 HRS Biomarker Data
#####################

# Set path to 2014 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/BIOMK14BL"))

# Load biomarker data
bio_2014 <- read.SAScii("BIOMK14BL.da", "BIOMK14BL.sas")

# Add unique ID and select relevant variables
bio_2014_trim <- bio_2014 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(
    unique_id,
    A1C = OA1C_ADJ,
    HDL = OHDL_ADJ,
    CRP = OCRP_ADJ
  ) %>%
  mutate(survey_year = 2014)

# Preview 2014 Biomarker Data
glimpse(bio_2014_trim)

```

---

### Load 2012 HRS Biomarker Data

```{r}

#####################
# Load 2012 HRS Biomarker Data
#####################

# Set path to 2012 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/biomkr12"))

# Load biomarker data
bio_2012 <- read.SAScii("BIOMK12BL_R.da", "BIOMK12BL_R.sas")

# Add unique ID and select relevant variables
bio_2012_trim <- bio_2012 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(
    unique_id,
    A1C = NA1C_ADJ,
    HDL = NHDL_ADJ,
    CRP = NCRP_ADJ
  ) %>%
  mutate(survey_year = 2012)

# Preview 2012 Biomarker Data
glimpse(bio_2012_trim)

```

---

### Load 2010 HRS Biomarker Data

```{r}

#####################
# Load 2010 HRS Biomarker Data
#####################

# Set path to 2010 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/biomkr10"))

# Load biomarker data
bio_2010 <- read.SAScii("BIOMK10BL_R.da", "BIOMK10BL_R.sas")

# Add unique ID and select variables of interest
bio_2010_trim <- bio_2010 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(
    unique_id,
    A1C = MA1C_ADJ,
    HDL = MHDL_ADJ,
    CRP = MCRP_ADJ
  ) %>%
  mutate(survey_year = 2010)

# Preview 2010 Biomarker Data
glimpse(bio_2010_trim)

```

---

### Load 2008 HRS Biomarker Data

```{r}

#####################
# Load 2008 HRS Biomarker Data
#####################

# Set path to 2008 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/biomkr08"))

# Load biomarker data
bio_2008 <- read.SAScii("BIOMK08BL_R.da", "BIOMK08BL_R.sas")

# Add unique ID and select variables of interest
bio_2008_trim <- bio_2008 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(
    unique_id,
    A1C = LA1CBIOS,
    HDL = LHDLBIOS_ADJ,
    CRP = LCRP_ADJ
  ) %>%
  mutate(survey_year = 2008)

# Preview 2008 Biomarker Data
glimpse(bio_2008_trim)

```

---

### Load 2006 HRS Biomarker Data

```{r}

#####################
# Load 2006 HRS Biomarker Data
#####################

# Set path to 2006 HRS Biomarker folder
setwd(file.path(path, "Biomarker Data/biomkr06"))

# Load biomarker data
bio_2006 <- read.SAScii("BIOMK06BL_R.da", "BIOMK06BL_R.sas")

# Add unique ID and select variables of interest
bio_2006_trim <- bio_2006 %>%
  mutate(unique_id = paste(HHID, PN, sep = "_")) %>%
  select(unique_id, A1C = KA1C_ADJ, HDL = KHDL_ADJ, CRP = KCRP_ADJ) %>%
  mutate(survey_year = 2006)

# Preview 2006 Biomarker Data
glimpse(bio_2006_trim)

```

---

### Combine 2006-2016 HRS Biomarker Data

```{r}

full_bio = rbind(bio_2016_trim, bio_2014_trim, bio_2012_trim, bio_2010_trim, bio_2008_trim, bio_2006_trim)

#Preview Combined Biomarker Data
glimpse(full_bio)

```

### Combine Full Biomarker and Core + Tracker Data

```{r}

# Combine Biomarker and Core + Tracker Data
full_core <- full_core %>% mutate(id_year = paste(unique_id, survey_year, sep = "_"))
full_bio  <- full_bio  %>% mutate(id_year = paste(unique_id, survey_year, sep = "_"))

full_data <- full_core %>%
  left_join(
    full_bio %>% select(id_year, A1C, HDL, CRP),
    by = "id_year"
  )

glimpse(full_data)

```

---

### Handle HRS special codes + set categorical variables as factors

```{r}

full_data <- full_data %>%
  mutate(
    # Pulse: recode non-substantive values and biologically implausible values to NA
    across(
      c(pulse1, pulse2, pulse3),
      ~ if_else(.x %in% c(998, 999) | .x >= 200, NA_real_, .x)
    ),

    # Waist: values >= 99 are out-of-range and/or non-substantive (includes 993/998/999)
    waist = if_else(waist >= 99, NA_real_, waist),

    # Height: 98 (DK) and 99 (RF)
    height = if_else(height >= 98, NA_real_, height),

    # Race: recode non-substantive values, retain masked "Other" category (97)
    race = if_else(race %in% c(98, 99), NA_real_, race),

    # Ethnicity: recode non-substantive values, retain masked "Other Hispanic" category (7)
    ethnicity = if_else(ethnicity %in% c(8, 9), NA_real_, ethnicity),

    # Convert categorical variables to factors (after recodes)
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    race = factor(race, levels = c(1, 2, 97), labels = c("White", "Black", "Other")),
    ethnicity = factor(ethnicity, levels = c(1, 7), labels = c("Mexican American", "Other Hispanic"))
  )

# Preview
glimpse(full_data)

```

### (Optional) Save / Reload Combined Dataset Before Allostatic Load Calculations

Loading raw HRS `.da` files can be time-consuming. To avoid re-running earlier data import and cleaning steps, you may wish to save the combined dataset at this stage and reload it directly in future sessions.

```{r}

# (Optional) Save combined dataset before AL calculations
# saveRDS(full_data, file.path(path, "hrs_combined_pre_AL.rds"))

# To reload in a future session without re-running raw imports:
# full_data <- readRDS(file.path(path, "hrs_combined_pre_AL.rds"))

```

---

## Part 2 – Derive Biomarker Risk Indicators & Compute Allostatic Load

Now that we’ve combined multiple HRS waves into a single dataset, we’re ready to compute risk thresholds and calculate our primary outcome: allostatic (over)load. This section includes:

- Computing Waist-to-Height Ratio (`WHtR`) from `waist` and `height`
- Creating `biomarker_count`, the number of non-missing biomarker values per observation (0–5)
- Computing wave-specific weighted quartile thresholds for each biomarker (by `survey_year`)
- Creating risk indicators for `CRP`, `A1C`, `HDL`, `pulse`, and `WHtR`
- Summing individual risk scores into a total Allostatic Load score (`al_score`, 0–5)

*Note:* This script uses HRS sampling weights to compute weighted percentile thresholds. It does not perform weighted inferential analyses (e.g., standard errors or hypothesis tests). Users conducting population-level inference should incorporate survey design variables (`STRATUM`, `SECU`) alongside the appropriate weights.


```{r}

#####################
# Allostatic Load Calculation
#####################

# Helper function for weighted quantiles:
# Returns NA if a survey year has no valid observations or weights.
wtd_q_safe <- function(x, w, p) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (sum(ok) == 0) return(NA_real_)
  as.numeric(Hmisc::wtd.quantile(x[ok], weights = w[ok], probs = p, na.rm = TRUE))
}

# Add derived variables: WHtR and age
full_data <- full_data %>%
  mutate(
    WHtR = waist / height,
    age = survey_year - birth_year
  )

# Create function to calculate pulse_clean (mean of last two heartbeats)
get_pulse <- function(pulse_dat) {
  if ((is.na(pulse_dat[2]) + is.na(pulse_dat[3])) == 0) {
    return(mean(c(pulse_dat[2], pulse_dat[3])))
  } else {
    return(mean(pulse_dat, na.rm = TRUE))
  }
}

full_data$pulse_clean <- apply(
  full_data %>% select(pulse1, pulse2, pulse3),
  MARGIN = 1,
  FUN = get_pulse
)

# Compute weighted thresholds within each survey year
thresholds <- full_data %>%
  group_by(survey_year) %>%
  summarise(
    WHtR_risk_threshold  = wtd_q_safe(WHtR,        WT_PM,  0.75),
    A1C_risk_threshold   = wtd_q_safe(A1C,         WT_BIO, 0.75),
    HDL_risk_threshold   = wtd_q_safe(HDL,         WT_BIO, 0.25),
    CRP_risk_threshold   = wtd_q_safe(CRP,         WT_BIO, 0.75),
    pulse_risk_threshold = wtd_q_safe(pulse_clean, WT_PM,  0.75),
    .groups = "drop"
  )

# Join thresholds back onto full_data
full_data <- full_data %>%
  left_join(thresholds, by = "survey_year")

# Calculate risk scores (set NA if biomarker or threshold missing)
full_data <- full_data %>%
  mutate(
    WHtR_risk_score  = ifelse(!is.na(WHtR)        & !is.na(WHtR_risk_threshold),  as.integer(WHtR >= WHtR_risk_threshold),  NA_integer_),
    A1C_risk_score   = ifelse(!is.na(A1C)         & !is.na(A1C_risk_threshold),   as.integer(A1C  >= A1C_risk_threshold),   NA_integer_),
    HDL_risk_score   = ifelse(!is.na(HDL)         & !is.na(HDL_risk_threshold),   as.integer(HDL  <= HDL_risk_threshold),   NA_integer_),
    CRP_risk_score   = ifelse(!is.na(CRP)         & !is.na(CRP_risk_threshold),   as.integer(CRP  >= CRP_risk_threshold),   NA_integer_),
    pulse_risk_score = ifelse(!is.na(pulse_clean) & !is.na(pulse_risk_threshold), as.integer(pulse_clean >= pulse_risk_threshold), NA_integer_)
  )

# Compute allostatic load score (NA if any component missing)
full_data <- full_data %>%
  mutate(
    al_score = ifelse(
      rowSums(is.na(select(., WHtR_risk_score, A1C_risk_score,
                           HDL_risk_score, CRP_risk_score, pulse_risk_score))) > 0,
      NA_integer_,
      as.integer(
        rowSums(select(., WHtR_risk_score, A1C_risk_score,
                       HDL_risk_score, CRP_risk_score, pulse_risk_score),
                na.rm = TRUE)
      )
    )
  )

# Add biomarker count (non-missing risk scores)
full_data <- full_data %>%
  mutate(
    biomarker_count = as.integer(
      rowSums(!is.na(select(., WHtR_risk_score, CRP_risk_score,
                            pulse_risk_score, HDL_risk_score, A1C_risk_score)))
    )
  )

# Trim to final dataset
full_trim <- full_data %>%
  select(
    id_year, unique_id, survey_year, age, sex, birth_year, race, ethnicity,
    WHtR, A1C, HDL, CRP, pulse_clean,
    WHtR_risk_threshold, A1C_risk_threshold, HDL_risk_threshold, CRP_risk_threshold, pulse_risk_threshold,
    WHtR_risk_score, A1C_risk_score, HDL_risk_score, CRP_risk_score, pulse_risk_score,
    al_score, biomarker_count, WT_RESP, WT_PM, WT_BIO, WT_LBQ, STRATUM, SECU
  )

glimpse(full_trim)
```

---

### Save and Export Prepared HRS Data

Export the final cleaned dataset in your preferred format. Below are examples using both `.csv` and `.rds` formats. 

```{r}

# Example: export as CSV (for use in Excel, SPSS, etc.)
# write.csv(full_trim, file.path(path, "hrs_allostatic_load_data.csv"), row.names = FALSE)

# Example: export as RDS (recommended for preserving datatypes and factors in R)
# saveRDS(full_trim, file.path(path, "hrs_allostatic_load_data.rds"))

```


---

### Using HRS Weights in Analysis

This dataset includes HRS sampling weights (`WT_RESP`, `WT_PM`, `WT_BIO`, `WT_LBQ`) and survey design variables (`SECU`, `STRATUM`) to support analyses that account for the complex survey design.  

A few general guidelines:  

- Use sample weights when estimating population statistics (means, proportions, percentiles, regression models, etc.) intended to represent the U.S. population.  
- Use both weights and design variables (`STRATUM`, `SECU`) when calculating standard errors, confidence intervals, or conducting hypothesis tests. These analyses typically use survey-design methods (e.g., the survey package in R).  
- Use the weight that corresponds to the subsample from which the variable was collected::  

  - Core interview variables (e.g., sex, birth_year, race, ethnicity) → `WT_RESP` 
  - Physical measures (e.g., pulse, waist, height) → `WT_PM`  
  - Biomarker variables (e.g., A1C, HDL, CRP) → `WT_BIO`  
  - Leave-behind questionnaire variables → `WT_LBQ` (if used)  
  
This script uses weights to compute biomarker risk thresholds but does not perform weighted inference; users conducting inferential or population-level analyses should apply survey weights and design variables appropriately.  

For detailed guidance, see the [HRS Cross-Wave Tracker File and Sampling Weights documentation](https://hrsdata.isr.umich.edu/data-products/cross-wave-tracker-file):  
https://hrsdata.isr.umich.edu/data-products/cross-wave-tracker-file
